The 2018 heatwave and its implications on ozone induced damage on vegetation in a subarctic climate

### 2020-04-01 VegOzone.tex
First draft of large parts of Section 2.

\introduction  %% \introduction[modified heading if necessary]
\label{sec:intro}
\begin{itemize}
\item ozone in general
\item ozone induced damage in vegetation \emph{Contributor: Ane - maybe also ask Aud?}
  Ozone acts as an oxidative stress to plants. Its main action is imposed through reactions occuring in the cell walls and cell membranes of mesophyll cells inside the leaves.Ozone enters the leaves through stomata, that are kept open to allow for gas exchange between the leaf air spaces and the exterior air. When the stomata are closed, ozone does not enter.
\item present subarctic climate and conditions
  \begin{itemize}
  \item[$\rightarrow$] midnight sun and 23\,\unit{h} open stomata (biological details will be published in a component paper)
  \item[$\rightarrow$] ozone climatology and spring peak
  \end{itemize}
\item future climate and conditions
  \begin{itemize}
  \item[$\rightarrow$] increase in heatwaves and dry periods
  \item[$\rightarrow$] shift in start of growing season (into vicinity of ozone spring peak)
  \item[$\rightarrow$] increase in tropospheric background ozone
  \end{itemize}
\item meteorological situation 2018 vs 2019
\end{itemize}

@@

In our assessment on ozone damage on vegetation in the subarctic in relation to the 2018 heatwave, we focus on the NIBIO operated local site Svanhovd which lies close by the settlement of Svanvik in the Pasvik valley (Section~\ref{subsec:svanhovd}). Throughout the paper, we will use Svanvik synonymously for the atmospheric monitoring site at Svanhovd. The available agrameteorological monitoring data are described in Section~\ref{subsubsec:atmo_svanvik}. In Section~\ref{subsubsec:ozone_garden}, the ozone garden -- a plant-based ozone monitoring -- will be described in more detail. For evaluation of the ozone data taken at Svanhovd, we compare with observational data from other monitoring sites in northern Fennoscandia (Section~\ref{subsec:obs_data}). Reanalysis data is used as complement where observational data was not available or temporally sparse. The chosen data sets are described in Section~\ref{subsec:model_data}.

\subsection{NIBIO research station Svanhovd}
\label{subsec:svanhovd}
"Svanhovd is located beside the Pasvik river, in the middle of the wedge of Norwegian land separating Russia and Finland in the north. It is where the eastern Siberian taiga meets the western Boreal forest. [...] Svanhovd was originally established as a State Demonstration and Experimental Farm in the 1930's, in order to expand agriculture in Pasvik valley [...]" - \href{https://www.nibio.no/en/about-eng/addresses/northern-norway/svanhovd}{NIBIO webpage, accessed March 2020}.\\
The research station at Svanhovd comprises long-term observations of various agrameteorological variables, i.a. $2\,\unit{m}$ temperature, precipitation, global radiation, and soil moisture and temperature at different depths. For the purpose of relating observed damage on ozone sensitive plants (e.g. clover and tobacco) an ozone monitor had been installed during the growing seasons 2018/19). The relevant data are described in more detail Section~\ref{subsubsec:atmo_svanvik}.
Note that from 1986--1996, ozone monitoring has been conducted operationally at Svanhovd. Details about these historical ozone data will be given in Section~\ref{subsubsec:ebas}.\\
The locations of the atmospheric monitoring site as well as the ozone garden with respect to the main buildings of the NIBIO research station are marked in the aerial photography shown in Fig.~\ref{fig:svanhovd_research_station}. 
\emph{Comment: Invite co-authors from NILU and NIBIO?}

@@

\subsubsection{Atmospheric monitoring}
\label{subsubsec:atmo_svanvik}
Monitoring data of key variables in the growing seasons 2018/19 have been downloaded from \href{luftkvalitet.no} operated by NILU and are shown in Fig.~\ref{fig:data_svanvik_2018_2019}. The hatched areas mark times when no ozone data has been taken. Note, while the downtime during winter was planed, missing data in two weeks of July 2018 were due to problems in data acquisition. For longer time series: LandbruksMeteorologiskTjeneste (NIBIO)

@@

\emph{Comment: We refrain from doing a full cross-evaluation for these meteorological fields, since this has been done elsewhere. It's somewhat clear that this is a job of the operators in the monitoring network.} {\bf TODO: Find a citation for this.}.
\begin{itemize}
\item Incoming radiation \emph{Comment to Ane: Same plot that you've included in the Nordic ozone report?}
\item ...
\end{itemize}

\begin{figure*}[t]
  \includegraphics[width=12cm]{data_svanvik_2018_2019}
  \caption{Observational data from atmospheric monitoring at Svanhovd in 2018/19. The hatched areas indicate missing ozone monitoring data. (a) Ozone; (b) temperature; (c) daily accumulated precipitation; (d) hourly global radiation.}
  \label{fig:data_svanvik_2018_2019}
\end{figure*}

\subsubsection{Ozone garden}
\label{subsubsec:ozone_garden}
      {\bf TODO: Details about the ozone garden.} \emph{Author: Ane/Aud}

\subsection{Other observational data}
\label{subsec:obs_data}
For the purpose of this study, we are interested in past and present observations of surface ozone concentrations and meteorological conditions in northern Fennoscandia. This limits the available observational data to a handful of atmospheric monitoring sites. An overview over their geographical locations is given in Fig.~\ref{fig:station_map_fennoscandia}.

@@

\subsubsection{Surface ozone observations}
\label{subsubsec:ebas}
All surface ozone observations used in this study are available from the EBAS database. In Table~\ref{tab:ebas_obs}, we give an overview over the exact locations and time range of operational data taking.

@@

\subsection{Reanalysis data}
\label{subsec:model_data}
Different reanalysis products have been used to evaluate and complement the observational data.

\subsubsection{Ozone}
\label{subsubsec:ozone_rea}
There are two global reanalysis products of ECMWF available that include also ozone. In addition, higher resolution surface ozone reanalysis for Europe is available from Copernicus service for regional air quality control. The details are given in Tab.~\ref{tab:ozone_rea}. The covered time periods differ but are sufficient for our purpose of comparison of derived ozone climatologies.

@@

\subsubsection{Temperature and precipitation}
Precipitation/Temperature: CRU bias corrected rea

@@

\subsection{Derived climatologies}
\begin{itemize}
\item temperature
\item precipitation
\item ozone
\end{itemize}

@@

\begin{figure}[t]
  \includegraphics[width=8.3cm]{ozone_climatology_fenoscandic_obs_spline}
  \caption{Derivation of ozone climatologies. Density distributions are shown together with multi-annual mean of daily ozone concentrations \chem{[O_3]}. Splines have been fitted though daily mean and max \chem{[O_3]}. The climatology of daily mean ozone is shown together with $1\,\sigma$ standard deviation and standard error of the mean. Since \chem{[O_3]} are strongly correlated for sites at Jergul/Karasjok (NOR), Esrange (SWE), and Pallas (FIN) (see Fig.~\ref{fig:density_distribution}), data of these have been averaged together to derive a climatology for northern Fennoscandia. (a) Northern Fennoscandia; (b) Svanvik}
  \label{fig:ozone_climatology_fenoscandic_obs_spline}
\end{figure}

@@

\begin{figure}[t]
  \includegraphics[width=8.3cm]{plot_temperature_anomalies_svanvik}
  \caption{Time series over the past 6 years of (a) temperature observed at Svanhovd and climatology derived from CRU bias-corrected reanalysis; (b) Student's t-test of temperature. The summer of 2018 has been significantly warmer on the $2-4\,\sigma$ level. The summer of 2019 was fairly average but on the colder side.}
  \label{fig:plot_temperature_anomalies_svanvik}
\end{figure}

\begin{figure}[t]
  \includegraphics[width=8.3cm]{plot_precipitation_anomalies_svanvik}
  \caption{Time series over the past 6 years of (a) precipitation observed at Svanhovd and climatology derived from CRU bias-corrected reanalysis; (b) Student's t-test of precipitation.}
  \label{fig:plot_precipitation_anomalies_svanvik}
\end{figure}

\begin{figure}[t]
  \includegraphics[width=8.3cm]{global_rad_signific}
  \caption{Global radiation at Svanvik. Percentage of days above/below $\pm 1\,\sigma$ from climatological mean displayed for each month. (a) 2018; (b) 2019. Negative deviations from climatology are shown as negative percentaged. The annual positive/negative deviation are indicated in the respective corners (right upper/ right lower).}
  \label{fig:global_rad_signific}
\end{figure}

@@

\begin{figure}[t]
  \includegraphics[width=8.3cm]{global_rad_clim}
  \caption{Observed global radition at Svanvik. The grey lines indicate the $1\,\sigma$ level deducted from the climatology. (a) Climatology (hourly maxima, mean, and minima); deviation from climatology (b) 2018; (c) 2019.}
  \label{fig:global_rad_clim}
\end{figure}

>>>>

\introduction  %% \introduction[modified heading if necessary]
\label{sec:intro}
\begin{itemize}
\item ozone in general
\item ozone induced damage in vegetation \emph{Contributor: Ane - maybe also ask Aud?}
  Ozone acts as an oxidative stress to plants. Its main action is imposed through reactions occurring in the cell walls and cell membranes of mesophyll cells inside the leaves.Ozone enters the leaves through stomata, that are kept open to allow for gas exchange between the leaf air spaces and the exterior air. When the stomata are closed, ozone does not enter.
\item present subarctic climate and conditions
  \begin{itemize}
  \item[$\rightarrow$] midnight sun and 23\,\unit{h} open stomata (biological details will be published in a component paper)
  \item[$\rightarrow$] ozone climatology and spring peak
  \end{itemize}
\item future climate and conditions
  \begin{itemize}
  \item[$\rightarrow$] increase in heatwaves and dry periods
  \item[$\rightarrow$] shift in start of growing season (into vicinity of ozone spring peak)
  \item[$\rightarrow$] increase in tropospheric background ozone
  \end{itemize}
\item meteorological situation 2018 vs 2019
\end{itemize}

@@

In our assessment on ozone damage on vegetation in the subarctic in relation to the 2018 heatwave, we focus on the NIBIO operated local site Svanhovd which lies close by the settlement of Svanvik in the Pasvik valley (Section~\ref{subsec:svanhovd}). Throughout the paper, we will use Svanvik synonymously for the atmospheric monitoring site at Svanhovd. The available meteorological monitoring data are described in Section~\ref{subsubsec:atmo_svanvik}. In Section~\ref{subsubsec:ozone_garden}, the ozone garden -- a plant-based ozone monitoring -- will be described in more detail. For evaluation of the ozone data taken at Svanhovd, we compare with observational data from other monitoring sites in northern Fennoscandia and use as well as reanalysis products, i.a., as complement where observational data was not available or spatially sparse, (Section~\ref{subsec:obs_data}).

\subsection{NIBIO research station Svanhovd}
\label{subsec:svanhovd}
"Svanhovd is located beside the Pasvik river, in the middle of the wedge of Norwegian land separating Russia and Finland in the north. It is where the eastern Siberian taiga meets the western Boreal forest. [...] Svanhovd was originally established as a State Demonstration and Experimental Farm in the 1930's, in order to expand agriculture in Pasvik valley [...]" - \href{https://www.nibio.no/en/about-eng/addresses/northern-norway/svanhovd}{NIBIO webpage, accessed March 2020}.\\
The research station at Svanhovd comprises long-term observations of various agrameteorological variables, i.a. $2\,\unit{m}$ temperature, precipitation, global irradiance, and soil moisture and temperature at different depths.

For the purpose of relating observed damage on ozone sensitive plants (e.g. clover and tobacco) an ozone monitor had been installed during the growing seasons 2018/19. The relevant data are described in more detail Section~\ref{subsubsec:atmo_svanvik}.

The locations of the atmospheric monitoring site as well as the ozone garden with respect to the main buildings of the NIBIO research station are marked in the aerial photography shown in Fig.~\ref{fig:svanhovd_research_station}. 
\emph{Comment: Invite co-authors from NILU and NIBIO?}
{\bf TODO: Some more general information about the research station? E.g., what NIBIO us currently researching there, that it also hosts a conference site, and national park center.}

@@

\subsubsection{Atmospheric monitoring}
\label{subsubsec:atmo_svanvik}
For the growing seasons 2018/19, monitoring data of key variables (temperature, precipitation, and ozone) have been downloaded from \href{luftkvalitet.no}{luftkvalitet.no} which is operated by NILU.

All agrameterological variables including temperature and precipitation are available from approximately 1992 to present day from LandbruksMeteorologisk Tjeneste provided by NIBIO (\href{https://lmt.nibio.no/}{lmt.nibio.no}, note the station name here is Pasvik).

@@

\emph{Comment: We refrain from doing a full cross-evaluation for these meteorological fields, since this has been done elsewhere. It's somewhat clear that this is a job of the operators in the monitoring network.} {\bf TODO: Find a citation for this.}.

Relevant data for the growing season 2018/19 are shown in Fig.~\ref{fig:data_svanvik_2018_2019}. The hatched areas mark times when no ozone data has been taken. Note, while the downtime during winter was planed, missing data in two weeks of July 2018 were due to problems in data acquisition.

Ozone concentrations \chem{[O_3]} measured in $2\,\unit{m}$ height above ground are averaged hourly. As can be seen in Fig.~\ref{fig:data_svanvik_2018_2019}a), \chem{[O_3]} peaks in spring (April/May) and reaches its minimum in late summer (July/August). The spring peak has not been captured in 2019, for data acquisition started later than in 2018. In 2018, high summer ozone concentrations ($\chem{[O_3]}> 40\,\unit{ppb}$) were observed in July. A part of these peak values may, however, coincident with the time period of missing data. In the summer of 2019, ozone concentrations stayed below the threshold of $40\,\unit{ppb}$.

Hourly averaged $2\,\unit{m}$ temperatures below $-20\,\unit{^\circ C}$ were observed in winter (January/February), while high temperatures (above $20\,\unit{^\circ C}$) occurred in summer (July), though more regularly in 2018 than in 2019 (Fig.~\ref{fig:data_svanvik_2018_2019}b). In 2018, temperature regularly rose above freezing only in May, while in 2019 this occurred already early in March/April. 2019 saw a very cold November with temperatures regularly below $-10\unit{^\circ C}$.

More rain events with accumulated daily precipitation $\sum_d \mathrm{Precip}$ above $10\,\unit{mm}$ occurred in the summer of 2018 compared to 2019 (Fig.~\ref{fig:data_svanvik_2018_2019}c). $\sum_d \mathrm{Precip} \ge 20\,\unit{mm}$ was only breached twice in 2018 and once in 2019.

In 2018, global irradiance $Q_0$ (Fig.~\ref{fig:data_svanvik_2018_2019}d) displays more sequentially high values in spring (May) and summer (July) compared to 2019, while June 2019 saw more consecutive high irradiance than 2018.

\begin{figure*}[t]
  \includegraphics[width=12cm]{data_svanvik_2018_2019}
  \caption{Observational data from atmospheric monitoring at Svanhovd in 2018/19. The hatched areas indicate missing ozone monitoring data. (a) Hourly averaged ozone concentration; (b) hourly averaged temperature; (c) daily accumulated precipitation; (d) hourly averaged global irradiance.}
  \label{fig:data_svanvik_2018_2019}
\end{figure*}

\subsubsection{Ozone garden}
\label{subsubsec:ozone_garden}
      {\bf TODO: Details about the ozone garden.} \emph{Author: Ane/Aud -- pictures are welcome, just upload them on overleaf or send them.}

\subsection{Other data}
\label{subsec:obs_data}
To determine the significance of the conditions in 2018, we are interested not only in present (2018/19) surface ozone concentrations at Svanvik but also in long term observations and a climatology derived thereof. While data for the period 1986--1996 exists for Svanvik, no ozone measurements have been conducted at Svanhovd for the past $14\,\unit{years}$. Hence, we also have to have a look at ozone observations at other sites in northern Fennoscandia. In regard to ozone, available observational data are limited to a handful of atmospheric monitoring sites. An overview over their geographical locations is given in Fig.~\ref{fig:station_map_fennoscandia}. The data is descibed in Section~\ref{subsubsec:ebas}. 
As additional complement to observational data which was not available at all times or spatially sparse, we also look at verious ozone reanalysis products (Section~\ref{subsubsec:ozone_rea}).

@@

\subsubsection{Surface ozone observations}
\label{subsubsec:ebas}
All surface ozone observations used in this study are available from the EBAS database operated by NILU (\href{http://ebas.nilu.no/}{ebas.nilu.no}). In Table~\ref{tab:ebas_obs}, we give an overview over the exact locations and time range of operational data taking.
Note that from 1986--1996, ozone monitoring has been conducted operationally by NILU at Svanhovd.

@@

\subsubsection{Ozone reanalysis}
\label{subsubsec:ozone_rea}

Different reanalysis products have been used to evaluate and complement the observational data.

There are two global reanalysis products of ECMWF available that include also ozone. In addition, higher resolution surface ozone reanalysis for Europe is available from Copernicus service for regional air quality control. The details are given in Tab.~\ref{tab:ozone_rea}. The covered time periods differ but are sufficient for our purpose of comparison of derived ozone climatologies.

@@

@@

\subsection{Derived climatologies}
\label{subsec:climatologies}

\subsubsection{Temperature and precipitation}
\label{subsubsec:clim_temp_prec}
\begin{figure}[t]
  \includegraphics[width=8.3cm]{climatediagram}
  \caption{Climate diagram following Walter and Lieth. The diagram is based on climatological data for Svanvik/Pasvik (1992--2017) downloaded from LandbruksMeteorologiske Tjeneste. Monthly average temperatures (red line) are displayed with standard error of mean as error band. Average monthly accumulated precipitation (blue bars) is shown with standard deviation of mean as error bars. Months with average temperatures below freezing are denoted with a star.}
  \label{fig:climatediagram}
\end{figure}

\subsubsection{Ozone}
\label{subsubsec:clim_ozone}

@@

\begin{figure}[t]
  \includegraphics[width=8.3cm]{ozone_climatology_fenoscandic_obs_spline}
  \caption{Derivation of ozone climatologies. Density distributions are shown together with multi-annual mean of daily ozone concentrations \chem{[O_3]}. Splines have been fitted though daily mean and max \chem{[O_3]}. The climatology of daily mean ozone is shown together with $1\,\sigma$ standard deviation and standard error of the mean. Since \chem{[O_3]} are strongly correlated for sites at Jergul/Karasjok (NOR), Esrange (SWE), and Pallas (FIN) (see Fig.~\ref{fig:density_distribution}), data of these have been averaged together to derive a climatology for northern Fennoscandia. (a) Northern Fennoscandia; (b) Svanvik.}
  \label{fig:ozone_climatology_fenoscandic_obs_spline}
\end{figure}

@@

\begin{figure}[t]
  \includegraphics[width=8.3cm]{plot_temperature_anomalies_svanvik}
  \caption{{\bf TODO: Replace by analysis based on Svanvik observed climatology.} Time series over the past 6 years of (a) temperature observed at Svanhovd and climatology derived from CRU bias-corrected reanalysis; (b) Student's t-test of temperature. The summer of 2018 has been significantly warmer on the $2-4\,\sigma$ level. The summer of 2019 was fairly average but on the colder side.}
  \label{fig:plot_temperature_anomalies_svanvik}
\end{figure}

\begin{figure}[t]
  \includegraphics[width=8.3cm]{plot_precipitation_anomalies_svanvik}
  \caption{{\bf TODO: Replace by analysis based on Svanvik observed climatology.} Time series over the past 6 years of (a) precipitation observed at Svanhovd and climatology derived from CRU bias-corrected reanalysis; (b) Student's t-test of precipitation.}
  \label{fig:plot_precipitation_anomalies_svanvik}
\end{figure}

\begin{figure}[t]
  \includegraphics[width=8.3cm]{global_rad_signific}
  \caption{Global irradiance at Svanvik. Percentage of days above/below $\pm 1\,\sigma$ from climatological mean displayed for each month. (a) 2018; (b) 2019. Negative deviations from climatology are shown as negative percentage. The annual positive/negative deviation are indicated in the respective corners (right upper/ right lower).}
  \label{fig:global_rad_signific}
\end{figure}

@@

\begin{figure}[t]
  \includegraphics[width=8.3cm]{global_rad_clim}
  \caption{Observed global irradiance at Svanvik. The grey lines indicate the $1\,\sigma$ level deducted from the climatology. (a) Climatology (hourly maxima, mean, and minima); deviation from climatology (b) 2018; (c) 2019.}
  \label{fig:global_rad_clim}
\end{figure}
